A meta-analysis of fungicide effects on mungbean powdery mildew

### 2021-05-20 04_Severity_MetaAnalysis.Rmd
Remove irrelavant calculations and add some words

### Inspect PM severity for normality
Given the meta-analysis assumes powdery mildew severity observations to be a univariate normal distubution $Y_i = N(\mu , \sigma)$, with a mean $\mu$ and standard deviation $\sigma$. We need to inspect severity for a normality.
```{r sev_hist}
hist(PM_dat_D$PM_final_severity)
summary(PM_dat_D$PM_final_severity)
```
We can see here that powdery mildew severity observations are skewed towards the top of the scale, eight and nine in this case.

Traditionally disease severity is `log` transformed due to it being more likely to be skewed to the lower disease severity scores. In this case if we `log()` transformed the response it would increase the skew towards higher disease severity.

We can improve the normality by transforming the effect size, disease severity, by squaring.
```{r}
hist(PM_dat_D$PM_final_severity^2)
summary(PM_dat_D$PM_final_severity^2)
```

### Transform severity response
As variance is proportional to the mean we square the variance as well.
```{r square_severity}
PM_dat_D$yi_square <- PM_dat_D$PM_final_severity ^2
PM_dat_D$vi_square <- (PM_dat_D$vi +1)^2
```
_Note: Here I added 1 to the variance before it was squared. This is because squaring numbers below one makes the number smaller or remain zero. This does not change the meta-analysis mean estimates but does increase heterogeneity and tao._

@@

```{r}
# Disease response ratio
# create column giving the respective disease severity in the control 
PM_dat_D$yi_C <- unlist(apply(PM_dat_D, 1, function(x1){

   mean(PM_dat_D[PM_dat_D$trial == x1[["trial"]] &
               PM_dat_D$spray_management == "control", "PM_final_severity"])
}))

# Calculate response ratio
PM_dat_D$yi_RR <- PM_dat_D$PM_final_severity/PM_dat_D$yi_C
hist(PM_dat_D[PM_dat_D$spray_management != "control", "yi_RR"])

# Log response ratio 
PM_dat_D$yi_LR <- log(PM_dat_D$yi_RR)
hist(PM_dat_D[PM_dat_D$spray_management != "control", "yi_LR"])

# Percent severity conversion
# define ordinal conversion from scale to proportion
s1 <- 0
s2 <- 0.33*0.5
s3 <- 0.5*0.87
s4 <- 0.66*0.75
s5 <- 0.66*0.87
s6 <- 0.66*1
s7 <- 0.75*1
s8 <- 1*0.87
s9 <- 1*1
x <- c(s1, s2, s3, s4, s5, s6, s7, s8, s9)
y <- c(1:9)
a = data.frame(x,y)
plot(a)

# fit a curve using a gam with spline value 0.2
g1 <- gam(x ~ s(y,spar = 0.1))

@@

# predict values for dataset
new_sev <- predict(object =  g1, 
                   newdata = data.frame(y = PM_dat_D[,"PM_final_severity"])
                   )
# one value is a negative, this should be a zero
new_sev[12] <- 0

# inspect distribution
hist(new_sev)
PM_dat_D$yi_percent <- new_sev

@@

# compare untransformed variance to transformed variance
summary(PM_dat_D$vi)
summary(PM_dat_D$vi_percent)

@@

PM_dat_D[PM_dat_D$yi_log == -Inf,"yi_log"] <- 0
PM_dat_D[PM_dat_D$vi_log == -Inf,"vi_log"] <- 0.001

# create column giving the respective disease severity in the control 
PM_dat_D$yi_Cp <- unlist(apply(PM_dat_D, 1, function(x1){

   mean(PM_dat_D[PM_dat_D$trial == x1[["trial"]] &
               PM_dat_D$spray_management == "control", "yi_percent"])
}))

# calculate response ratio
PM_dat_D$L_yi_p  <- log((PM_dat_D$yi_percent / PM_dat_D$yi_Cp) + 1)

# transform variance as per Paul et. al 2007
PM_dat_D <- 
   PM_dat_D %>%
   mutate(L_RR_v = vi *(1/replicates * (yi_percent ^ 2) + 1/replicates * (yi_Cp^2)))

>>>>

### Inspect PM severity for normality
Given the meta-analysis assumes powdery mildew severity observations to be a univariate normal distribution $Y_i = N(\mu , \sigma)$, with a mean $\mu$ and standard deviation $\sigma$. We need to inspect severity for a normality.
```{r sev_hist}
hist(PM_dat_D$PM_final_severity)
summary(PM_dat_D$PM_final_severity)
```
We can see here that powdery mildew severity observations are skewed towards the top of the scale, eight and nine in this case.

We can improve the normality by transforming the effect size, disease severity, by squaring.
```{r}
PM_dat_D$yi_square <- PM_dat_D$PM_final_severity ^2
PM_dat_D$vi_square <- (PM_dat_D$vi +1)^2
# Note: Here I added 1 to the variance before it was squared. This is because squaring numbers below one makes the number smaller or remain zero. This does not change the meta-analysis mean estimates but does increase heterogeneity and tao._

hist(PM_dat_D$yi_square)
summary(PM_dat_D$yi_square)
```

Another transformation which might be more suitable for a discrete response is either a `log()` or `logit()` transformation. 
However first we should convert the scale from 1 - 9 to between 0 and 1.

### Transform severity response

@@

```{r}
# Percent severity conversion
# define ordinal conversion from scale to proportion
s1 <- 0
s2 <- 0.33*0.5
s3 <- 0.5*0.87
s4 <- 0.66*0.75
s5 <- 0.66*0.87
s6 <- 0.66*1
s7 <- 0.75*1
s8 <- 1*0.87
s9 <- 1*1
x <- c(s1, s2, s3, s4, s5, s6, s7, s8, s9)
y <- c(1:9)
a = data.frame(x,y)
plot(a)
```

We can fit a gam curve between the 1 - 9 scale and the new scale between 0 and 1
```{r}
# fit a curve using a gam with spline value 0.2
g1 <- gam(x ~ s(y,spar = 0.1))

@@

# predict values for dataset
new_sev <- predict(object =  g1, 
                   newdata = data.frame(y = PM_dat_D[,"PM_final_severity"])
                   )
# Check fit
plot(x = new_sev, y = PM_dat_D[,"PM_final_severity"])
points(x,y,col = "red", pch = 16)
```

```{r}
# one value is a negative, this should be a zero
# new_sev[12] <- 0

# Allocate new variable percent severity
PM_dat_D$yi_percent <- new_sev

@@

@@

# Correct Inf to 0
PM_dat_D[PM_dat_D$yi_log == -Inf,"yi_log"] <- 0
PM_dat_D[PM_dat_D$vi_log == -Inf,"vi_log"] <- 0.001
